---
title: "Four ways to perform Granger Causality test with R"
output: github_document
author: Nicola Righetti
website: https::www.nicolarighetti.net 
---

Granger causality test is common practice in time series analysis, but the different ways it can be performed in R are poorly documented. I already published a function to perform the Toda-Yamamoto bivariate and multivariate versions of the Granger causality test (see at  [Toda-Yamamoto-Causality-Test](https://github.com/nicolarighetti/Toda-Yamamoto-Causality-Test/blob/main/README.md). In this brief document, I am going to focus on its standard version, suggesting five different approaches.

A time series X is said to Granger cause another time series Y if past values of X and Y predict Y significantly better than past values of Y alone (Granger, 1969, see  [here](https://github.com/nicolarighetti/Toda-Yamamoto-Causality-Test/blob/main/README.md)). We can test this hypothesis by using linear regression and a Wald test for linear restrictions. The Wald test compares the performance of a restricted model for Y, which excludes X, against an unrestricted model for Y, which includes X.

First off, let's create two time series. It's important to notice that the data are represented as time series using the base R *ts* function. Some functions documented here do not function properly if the data are not in this format.

```{r setup, include=FALSE}
set.seed(1234)

ts1 <- ts(rnorm(n=100))
ts2 <- ts(rnorm(n=100))
```

## First approach: lmtest::grangertest

The first way to perform a Granger causality test in R is to use the *grangertest* function provided by the [lmtest](https://cran.r-project.org/web/packages/lmtest/index.html) package. To test if the series *ts1* Granger-causes the series *t2* we can run a simple line of code, as follows. It turns out the *ts1* Granger-causes *ts2*.

```{r}
lmtest::grangertest(ts1, ts2, order=3, test="F")
```

While *grangertest* is a quick way to perform a Granger causal test, it is also quite rough, as it does not rely on any modelling of the series.

## Second approach: dynlm::dynlm + lmtest::waldtest

A second, more "manual" and thus flexible way to perform the test, is based on performing a Wald Test on two linear regression models. Linear regression models including lags of the variables can be easily fitted with the function *dynlm* of the homonymous library [dynlm](https://cran.r-project.org/web/packages/dynlm/index.html). Next, a Wald test can be performed by using the function *waldtest* of the already mentioned library *lmtest*.

```{r}
library(dynlm)
unrestricted_ts2 <- dynlm(ts2 ~ L(ts2, 1:3) + L(ts1, 1:3))
restricted_ts2 <- dynlm(ts2 ~ L(ts2, 1:3))

lmtest::waldtest(unrestricted_ts2, restricted_ts2, test="F")                  
```

The results are the same as above, in this case, but this approach has the advantage that the series can be modeled flexibly prior to testing Granger's causality. For instance, by adding a seasonality term or other variables when necessary.

Morevoer, this approach permits to specify the heteroscedasticity-consistent estimation (HC) or the heteroscedasticity and autocorrelation consistent estimation (HAC) of the covariance matrix of the coefficient estimates, also in the Newey West form, for instance by using the functions provided by the [sandwich](https://cran.r-project.org/web/packages/sandwich/index.html) package.

```{r}
lmtest::waldtest(unrestricted_ts2, restricted_ts2, test="F", vcov = sandwich::vcovHC) 
lmtest::waldtest(unrestricted_ts2, restricted_ts2, test="F", vcov = sandwich::vcovHAC) 
lmtest::waldtest(unrestricted_ts2, restricted_ts2, test="F", vcov = sandwich::NeweyWest) 
```

## Third approach: dynlm::dynlm + aod::wald.test

The third approach is a version of the latter. A Wald test can also be done using the *wald.test* function of the [aod](https://cran.r-project.org/web/packages/aod/index.html) package. It is trickier because you must manually specify the terms to be tested. However, this additional flexibility can prove useful in certain cases. By default, *wald.test* will calculate a Wald Chi-square test. The Chi-squared version of the test can be also computed with the *waldtest* function of *lmtest* by specifying the option test="Chisq". Similarly, the F or Chi-squared test can be performed by using the the function *linearHypothesis* in the [car](https://cran.r-project.org/web/packages/car/index.html) package.

```{r}
library(aod)
aod::wald.test(b=coef(unrestricted_ts2), 
               Sigma=vcov(unrestricted_ts2),
               Terms = 5:7,
               verbose = T)
```

## Fourth approach: vars::VAR + vars::causality

The fourth way to perform a Granger causality test in R is probably the most commonly used, and consists in using the package [vars](https://cran.r-project.org/web/packages/vars/index.html) to fit a VAR model (which is basically a system of linear regression models) and subsequently run the Granger test using the function *causality* provided by the same package.

```{r}
tsDat <- ts.union(ts1, ts2) 
tsVAR <- vars::VAR(tsDat, p = 3)
vars::causality(tsVAR, cause = "ts1")$Granger
```

The *causality* function has several interesting options. For example, it implements a bootstrapping procedure. It is also possible to specify the heteroscedasticity-consistent estimation of the covariance matrix of the coefficient estimates in the regression models (HC), but other types of estimation of the covariance matrix (e.g., HAC) are not supported.